---
title: "ESM 206 Lab 5"
author: "Lizzy Schattle"
date: "10/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


#### Lab 5 Objectives

- learn to parse dates with 'lubridate'
- get counts of observations with count()
- uncount() rows
- one and two sample t-test
- create heatmap with geom_tile()

```{r, include = FALSE}

# attach packages

library(tidyverse)
library(here)
library(janitor)

# read in data

lobster_abundance <- read_csv(here::here("data", "lobster_abundance.csv"), 
  na = "-99999") %>% 
  janitor::clean_names()

# this data bad, gotta clean
# need to change date from character, -99999 to NA, all caps to lower snake, count for each individual lobster



```

Use tidyr::uncount() function to convert our lobster data from frequency format to case format

```{r}
lobster_tidy <- lobster_abundance %>% 
  tidyr::uncount(lobster_count)
```

#### Exploratory Data Visualization

we'll only consider 'site' as our variable of interest

```{r}

ggplot(lobster_tidy, aes(x = site, y = size_mm)) +
  geom_jitter(aes(color = site),
              alpha = 0.5,
              width = 0.2)
# histogram

ggplot(lobster_tidy, aes(x = size_mm)) +
  geom_histogram(aes(fill = site)) +
  facet_wrap(~site, scales = "free")

# quantile-quantile plot to tell if the data have a linear relationship compared to a normal distribution

ggplot(lobster_tidy, aes(sample = size_mm)) +
  geom_qq() +
  facet_wrap(~site)

```


#### Convert 'date' column to class 'Date'

we'll use the 'lubridate' package to convert to date format, and then to help us easily parse month and year

```{r}
lobster_date <- lobster_tidy %>% 
  mutate(
    date_new = lubridate::mdy(date)
  )

# checked with class(lobster_date$date_new)
```

now parse year and month using lubridate::month() and lubridate::year()

```{r}
lobster_parse_date <- lobster_date %>% 
  mutate(
    obs_month = lubridate::month(date_new, label = TRUE),
    obs_year = lubridate::year(date_new)
  )
```

find counts of observed lobsters based on different heirarchical groupings:

First, let's count lobsters by year and month

```{r}
lobster_ym <- lobster_parse_date %>% 
  dplyr::count(obs_year, obs_month)

lobster_y <- lobster_parse_date %>% 
  dplyr::count(obs_year)

lobster_site <- lobster_parse_date %>% 
  dplyr::count(site)
```

To create a summary table that contains statistics other than counts by group, it's easier to use group_by() + n()

```{r}
lobster_summary <- lobster_parse_date %>% 
  group_by(site) %>% 
  summarize(
    mean_size = mean(size_mm, na.rm = TRUE),
    sd_size = sd(size_mm, na.rm = TRUE),
    lobster_number = n()
  )
```

#### Find Confidence Intervals

use t.test() function to find confidence intervals (for one sample) and perform t-tests to compare means of two samples

```{r}
ivee_lobsters <- lobster_tidy %>% 
  filter(site == "IVEE") %>% 
  pull(size_mm)

t.test(ivee_lobsters)
```

#### 2 sample t-test to compare means

is there a significant difference in lobster lengths at Naples and Mohawk reefs?

we've done our necessary exploratory analyses (sample size large, normal distribution for both samples) to determine that a 2-sample t-test for means comparison is appropriate

```{r}
napl_sample <- lobster_tidy %>% 
  filter(site == "NAPL") %>% 
  pull(size_mm)

mohk_sample <- lobster_tidy %>% 
  filter(site == "MOHK") %>% 
  pull(size_mm)

mn_ttest <- t.test(mohk_sample, napl_sample)

mn_ttest
```


There is a significant difference in lobster lengths between Naples and Mohawk Reef (t(df) = statistic, p < 0.001, alpha =0.05).

```{r}
# model format

lobster_mn <- lobster_tidy %>% 
  filter(site %in% c("NAPL", "MOHK"))

mn_ttest2 <- t.test(size_mm ~ site, data = lobster_mn)

mn_ttest2
```


in line variable referencing (?t.test) in r-markdown

here is the p-value: `r mn_ttest2$p.value`
here is the t-statistic: `r mn_ttest2$statistic`


```{r}
mn_ttest2$p.value
mn_ttest2$statistic


```

#### Now a heatmap!

```{r}

lobster_ys <- lobster_parse_date %>% 
  count(obs_year, site)

ggplot(data = lobster_ys, aes(x = obs_year, y = site)) +
  geom_tile(aes(fill = n))

```







